---
title: "Seasonality Analysis"
author: 'Rami Krispin'
date: "10/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.height=5, fig.width=8,
                      message=FALSE, warning=FALSE)
```

Time series seasonal analysis is one of the core foundations of time series analysis. Generally, a time series could either have a seasonal component or not. In addition, a series could have more than one seasonal pattern, for example:

* Daily series could have both day of the year and day of the week seasonal patterns
* Hourly series could have both hourly, daily, monthly seasonal patterns

In this section, we will focus on time series with a single seasonality pattern. In the following examples, we will use the US monthly consumption of natural gas.

One of the main tools for exploring seasonality is data visualization. Another simplistic but useful is summary statistics tables.

```{r}
library(tsibble)
library(dplyr)
library(lubridate)
library(plotly)
library(ggplot2)
library(viridis)
library(fabletools)
library(feasts)
```

#### Data

First, let's load the series and reformat it into `tsibble` object:

```{r}
naturalgas_path <- paste(rprojroot::find_rstudio_root_file(), "data", "NATURALGAS.csv", sep = "/")

us_gas <- read.csv(naturalgas_path, stringsAsFactors = FALSE) %>%
  setNames(c("date", "y")) %>%
  mutate(date = yearmonth(as.Date(date))) %>%
  as_tsibble(index = "date")


head(us_gas)
```


Just by eyeballing the series plot, you can identify the strong seasonal pattern: 

```{r }
us_gas %>% plot_ly(x = ~ as.Date(date),
                   y = ~ y,
                   type = "scatter",
                   mode = "line",
                   name = "US Natural Gas") %>%
  layout(title = "US Natural Gas Consumption (Not Seasonally Adjusted)",
         yaxis = list(title = "Billion Cubic Feet"),
         xaxis = list(title = paste("Source: U.S. Bureau of Transportation Statistics,", "<br>" ,
                                    "Natural Gas Consumption [NATURALGAS]", sep = "")),
         hovermode = "compare")

```


#### Decomposing the Series

We can leverage the `classical_decomposition` function to decompose the series to the trend, seasonal and irregular components:

```{r}
us_gas %>% 
  model(classical_decomposition(y)) %>%
  components() %>% 
  autoplot()
```


#### Summary Statistics

We will start with the summary statistics tables by calculating:

* Yearly average
* Monthly average


We will create time features with the `year` and `month` functions from the **lubridate** package that enables us to group by those features:

```{r}
us_gas <- us_gas %>%
  mutate(year = year(date),
         month = month(date, label = TRUE))

head(us_gas)
```

The year average could provide us insights about change in the trend direction. This information can also be observed from the decompose function:

```{r}
us_gas %>% 
  as.data.frame() %>%
  group_by(year) %>%
  summarise(avg = mean(y)) %>%
  plot_ly(x = ~ year,
          y = ~ avg,
          type = "scatter",
          mode = "line")
```

If a series has a strong trend it may shadow some of the series seasonality. In this case, you can either use short part of the series or detrend the series.

The next summary is by the main frequency units - the month of the year:

```{r}
us_gas %>% 
  as.data.frame() %>%
  group_by(month) %>%
  summarise(avg = mean(y))
```

#### Box-plot plots

A box plot is a useful method to visualize the distribution of the series observations by its frequency units:

```{r}
us_gas %>% 
  plot_ly(x = ~month, y = ~ y, 
          type = "box", 
          color = ~month, 
          colors = "Dark2",
          boxpoints = "all", 
          jitter = 0.3,
          pointpos = -1.8)
```

While we can see a clear seasonal pattern, the distribution of each month is fairly wide. This is mainly due to the year over year growth (or the series trend). We can reduce the growth impact by detrending the series and replot the seasonal box-plot. We will use the `classical_decomposition` and `model` functions from the **feasts** and **fabletools** packages, respectively:

```{r}
us_gas_comp <- us_gas %>% 
  model(classical_decomposition(y)) %>%
  components() 

us_gas <- us_gas %>% 
  left_join(us_gas_comp %>% select(date, trend)) %>%
  mutate(detrend = y - trend)
```

Let's replot the series monthly distribution, this time using the detrended series:

```{r}
us_gas %>% 
  plot_ly(x = ~month, y = ~ detrend, 
          type = "box", 
          color = ~month, 
          colors = "Dark2",
          boxpoints = "all", 
          jitter = 0.3,
          pointpos = -1.8)
```

After reducing the trend impact, the monthly distribution, as expected, is narrowing down, and we can see a clear seasonal pattern.

#### Seasonal Plot

The seasonal plot splits the series by its seasonal cycle (year) and compares changes in each cycle (or from year to year). In the following example, we will create a seasonal plot with plotly:
 
```{r}
p <- plot_ly() 

years <- unique(us_gas$year)

colors <- viridis::viridis(n = length(years))

for(i in seq_along(years)){
  y <- NULL
  
  y <- us_gas %>% 
    as.data.frame() %>%
    filter(year == years[i])
  
  p <- p %>% add_lines(x = y$month,
                       y = y$y,
                       line = list(color = colors[i]),
                       name = years[i])
  
}

p
```

**Note:** that we use a color scale to mark the years. This enables us to quickly observe the series' growth from year to year and the seasonal pattern. 

Similarly, we can split the series by its frequency units and plot each frequency unit against the frequency cycle (e.g., year):

```{r}
p <- plot_ly()

for(i in unique(us_gas$month)){
  m <- NULL
  
  m <- us_gas %>%
    as.data.frame() %>%
    filter(month == i)
  
  p <- p %>% 
    add_lines(x = m$year,
              y = m$y,
              name = i)
}

p
```


#### Heatmap

Heatmap is another visualization tool for reviewing and identifying a seasonal pattern in a series:

```{r}
us_gas_df <- us_gas %>%
  as.data.frame() %>%
  select(year, month, y) %>%
  tidyr::pivot_wider(names_from = year,
                     values_from = y)


plot_ly(x = names(us_gas_df)[-1],
        y = us_gas_df$month,
        z = as.matrix(us_gas_df[, -1]),
        colors = "Reds",
        xgap = 3,
        ygap = 3,
        type = "heatmap")
```


#### Kernel Density Plot

Likewise in the box-plot seasonal plot, we can use the kernel density plot to visualize the series frequency units distribution. A simple method is to use the **ggplot2** package `geom_density` function:

```{r}
ggplot(us_gas, aes(x = y)) +
geom_density(aes(fill = month)) +
ggtitle("US Gas - Kernel Density Estimates by Month") + facet_grid(rows = vars(as.factor(month)))
```

As before, the detrended series should have narrow distribution:

```{r}
ggplot(us_gas, aes(x = detrend)) +
geom_density(aes(fill = month)) +
ggtitle("US Gas - Kernel Density Estimates by Month") + facet_grid(rows = vars(as.factor(month)))
```

#### Seasonal Plots with the feasts Package

The **feasts** package provides some built-in functions for seasonal plots of `tsibble` objects based on the **ggplot2** package. The `gg_season` function creates a seasonal plot by the series frequency cycle units. By default, the function labels each line by its frequency cycle unit:

```{r}
us_gas %>%
gg_season(y, labels = "both") +
  ylab("Billion Cubic Feet") +
  xlab("Month") +
  ggtitle("US Natural Gas Consumption Seasonal Plot")
```

The `gg_subseries` split and plot each frequency unit aginst its frequency cycle unit:


```{r}
us_gas %>%
  gg_subseries(y) +
    ylab("Billion Cubic Feet") +
    xlab("Year") +
    ggtitle("US Natural Gas Consumption by Month")
```

#### Exercise

Create seasonal plots for the US Alcoholic Beverages Sales:

```{r}
us_abs_path <- paste(rprojroot::find_rstudio_root_file(), "data", "Alcoholic Beverages Sales.csv", sep = "/")

us_abs <- read.csv(us_abs_path, stringsAsFactors = FALSE) %>%
  setNames(c("date", "y")) %>%
  mutate(index = yearmonth(as.Date(date))) %>%
  select(-date) %>%
  as_tsibble(index = index)
```

