---
title: "Correlation Analysis"
author: 'Rami Krispin'
date: "10/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.height=5, fig.width=8, 
                      message=FALSE, warning=FALSE)
```


Correlation analysis of time series goes side by side with seasonal analysis. The goal of correlation analysis is to provide insights on the relationship between the series and its lags. This information can then be used to define and tune the parameters of forecasting models such as ARIMA, TSLM, etc.

Correlation analysis is mainly based on data visualization and statistical tools.  

```{r}
library(TSstudio)
library(plotly)
library(feasts)
library(tsibble)
```


#### Data

In the following examples, we will use the `AirPassengers` series again:

```{r}
data("AirPassengers")

ts_plot(AirPassengers,
        slider = TRUE)
```


#### The Auto Correlation Function

The Auto Correlation Function, or ACF, is the main tool for quantifying the level of correlation between a series and its lags. This method is fairly similar (both mathematically and logically) to the Pearson correlation coefficient but has time awareness:

$$r_{k} = \frac{\sum_{t = k+1}^{n-k}(x_{t-k} - \overline{x})(x_t-\overline{x})}{\sum_{t = 1}^{n}(x_t - \overline{x})^2}, ~ where$$


* $r_k$ is the ACF correlation coefficient of the series with its k lag
* $n$ is the number of observations of the series
* $x_t$ is the t observation of the series, and
* $\overline{x}$ the series and the mean


In R, the `acf` function from the **stats** package is the main tool for calculating the series AC (Auto Correlation). It supports only `ts` objects. We later see the `ACF` function from the **feasts** package, a wrapper of the `acf` function that supports `tsibble` objects. Let’s use the `acf` function to calculate the AC of the `AirPassengers` series:

```{r}
acf(AirPassengers)
```

#### The Partial Auto Correlation Function

The Partial Auto Correlation Function (PACF) is a conditional correlation between a series and lag k, given the impact of the lags 1 to k-1 on the series. Likewise the `pcf` function, the `pacf` function is the corresponding function for calculating the PAC (Partial Auto Correlation) of a series with its lags:

```{r}
pacf(AirPassengers)
```


The key applications of the ACF and PACF functions are:

* Quantify the relation of the series with its past lags
* Identify and tune the AR and MA process order for ARIMA models (e.g., AR, MA, ARMA, ARIMA, SARIMA)
* Identify seasonal patterns
* Residual analysis and white noise test


#### ACF and PACF Wrappers

The `ts_cor` function from provides an interactibe wrapper for the `acf` and `pacf` functions, plotting both correlations:

```{r }
ts_cor(AirPassengers, lag.max = 72)
```

By default, the function marked the seasonal lags in red. You can use the `seasonal_lags` argument to mark additional seasonal lags (in case exists):

```{r}
ts_cor(AirPassengers, lag.max = 72, seasonal_lags = 3)
```


#### Lags Plots

The lag plot is a common method for visualizing the correlation between a series and its lags. The `ts_lags` function from the **TSstudio** package provides this functionality:

```{r}
ts_lags(AirPassengers)
```

As the relationship between a series and its past lag looks more linear, the higher the correlation between the series and the lag. In the case of the `AirPassengers` series, you can see that the series has a strong linear relationship with the first and seasonal (lag 12) lags, as observed before with the `acf` function.


The `lags` arguments enable us to plot the series against specific lags. For example, we could plot the relationship of the series with its past three seasonal lags (e.g., 12, 24, and 36):

```{r}
ts_lags(AirPassengers, lags = c(12,24, 36))
```


#### Correlation Analysis with the feasts Package

So far, the tools we saw above supports `ts` objects. As we saw before, the ** feasts ** provides wrappers for the **stats** main functions for time series analysis for `tsibble` objects. First, let's convert the `AirPassengers` to a `tsibble` object:


```{r}
ap_tsibble <- AirPassengers %>% as_tsibble()



```

The `ACF` and `PACF` functions, as their names impay, provides a warappers for the `acf` and `pacf` functions. By default, unlike the origin functions it won't plot the output and we will have to add the `autoplot` function to plot the results:

```{r}
ap_tsibble %>% ACF(value, lag_max = 48) %>% autoplot()
ap_tsibble %>% PACF(value, lag_max = 48) %>% autoplot()
```

The `gg_lag` provides lag plots. The nice feature of this function that it colored the observations by their frequency units. By default, it used `geom` line, which I found a bit confusing, so we will set the `geom` to `point`:


```{r}
p <- ap_tsibble %>% gg_lag(value, geom="point")

p
```


We can make this plot interactive by using the `ggplotly` function from the **plotly** package:

```{r}
ggplotly(p)
```

